---
title: "Primers and probes calibration vignette"
author: "Edward Wallace"
date: "Feb 2019"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PrimerCalibration}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# Summary: calibrating primer sets from a real experimental test

This vignette shows how to use tidyqpcr functions to calibrate qPCR probes.

This is real qPCR data by Edward Wallace in Feb 2019, testing new RT-qPCR primer sets against _S. cerevisiae_ genes. We took exponential-phase total RNA previously extracted by Jamie Auxillos.

We tested 2-3 primer sets each for 7 genes:

* ECM38/ YLR299W (3 primer sets)
* FET5 / YFL041W (3 primer sets)
* GPT2/YKR067W
* ILV5/YLR355C
* NRD1/YNL251C (ordered 3rd primer set not tested today)
* RDL1/YOR285W
* TFS1/YLR178C

We started with two biological replicate RNA samples, treated with DNase and then split for a test sample with reverse transcriptase (RT) and negative control without reverse transcriptase (-RT). We also took a no template (NT) negative control. For each RT reaction we do serial 5x dilutions down to 125x to form a quantitative calibration curve.

The data were measured on a Roche LC480 instrument in a single 384-well plate. Quantification was performed in the Roche LightCycler software prior to using this program.


```{r setup,warning=FALSE,message=FALSE,echo=FALSE}
## knitr options for report generation
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, echo = TRUE, cache = FALSE,
  results = "show"
)
library(tidyverse)
library(cowplot)
library(tidyqpcr)

# set default theme for graphics
theme_set(theme_cowplot(font_size = 11) %+replace%
  theme(
    panel.border = element_rect(
      colour = "grey50",
      linetype = "solid", size = 0.5
    ),
    strip.background = element_blank()
  ))
```

# Set up experiment

## Label and plan plates



```{r label_plates,dependson="plate_functions"}
# Names of target genes
Names <- c("ECM38", "FET5", "GPT2", "ILV5", "NRD1", "RDL1", "TFS1")
# ORF ids of target genes
target_levels <- c("YLR299W", "YFL041W", "YKR067W", "YLR355C", "YNL251C", "YOR285W", "YLR178C")
# Repeats of gene names to account for testing multiple primer sets
Namesrep <- c(rep(Names[1:2], each = 3), rep(Names[3:7], each = 2))
# id numbers of multiple probesets (reflecting IDs as ordered)
target_id_levels <- paste(Namesrep,
  c(1, 2, 3, 1, 3, 4, 1, 4, 1, 4, 1, 2, 4, 5, 1, 5),
  sep = "_"
)


rowkey <- data.frame(
  well_row = LETTERS[1:16],
  Names = Namesrep,
  target_id = factor(target_id_levels, levels = target_id_levels)
)


plate1plan <-
  label_plate_rowcol(
    create_blank_plate(),
    rowkey,
    create_colkey_4diln_2ctrl_in24()
  ) %>%
  mutate(sample_id = paste(BioRep, dilution_nice, sep = "_"))
```

## Spot-check the plate plan 

Checks that for selected techrep/probe/dilution combinations, the plate contains the right number of replicates.

```{r print_techreps,results="show",echo=TRUE,cache=FALSE}
plate1plan %>%
  filter(TechRep == "1", target_id == target_id_levels[1], dilution_nice == "1x")

plate1plan %>%
  filter(TechRep == "2", target_id == target_id_levels[4])
```

## Display the plate plan

This can be printed out to facilitate loading the plate correctly.

```{r display_plates,fig.height=8,fig.width=12,dependson="label_plates"}
display_plate(plate1plan)
```


# Analyze Cq (quantification cycle count) data

## Load and summarize data

```{r load_plates,dependson="label_plates",results="show"}

# read my plates

plates <- read_tsv("../inst/extdata/Edward_qPCR_Nrd1_calibration_2019-02-02_Ct.txt",
  skip = 1
) %>%
  mutate(well = Pos, cq = Cp) %>%
  right_join(plate1plan)
```


```{r show_plates,dependson="load_plates",results="show"}
plates

summary(plates)
```

## Plot unnormalized data shows that -RT and NT controls are low

We detect no signal in NT (no template) negative control, and very weak signal in -RT (no reverse transcriptase) negative controls.

```{r plot_unnormalized,dependson="load_plates",fig.height=6,fig.width=9}

ggplot(data = plates) +
  geom_point(aes(x = target_id, y = cq, colour = dilution_nice, shape = prep_type),
    position = position_jitter(width = 0.2, height = 0)
  ) +
  labs(
    y = "Cycle count to threshold",
    title = "All reps, unnormalized"
  ) +
  facet_wrap(~BioRep) +
  panel_border() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```



## Dilution series is linear for all probes

Visual display of linearity of cq with log(dilution).

```{r plot_dilutions,dependson="load_plates",fig.height=11,fig.width=6}
ggplot(data = filter(plates, prep_type == "+RT"), aes(x = dilution, y = cq)) +
  geom_point() +
  stat_smooth(
    formula = y ~ x, method = "lm", se = FALSE,
    aes(colour = "fit", linetype = "fit")
  ) +
  stat_smooth(
    formula = y ~ 1 + offset(-x * log(10) / log(2)), method = "lm", se = FALSE,
    aes(colour = "theory", linetype = "theory")
  ) +
  scale_x_log10(breaks = 10^-(0:3)) +
  scale_y_continuous(breaks = seq(0, 30, 2)) +
  labs(
    y = "Cycle count to threshold",
    title = "All reps, unnormalized",
    colour = "Dilution", linetype = "Dilution"
  ) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

## Calculate primer efficiencies for all probes

Use regression to estimate linearity of cq with log(dilution), including the slope or efficiency.

```{r calibrate_dilutions,dependson="load_plates",results="show",echo=TRUE}
calculate_efficiency_bytargetid(plates)
```


## Dilution series for nice probes only shows linearity clearly

```{r plot_dilutions_nice,dependson="load_plates",fig.height=6,fig.width=4}
target_id_levels_nice <- c("ECM38_3", "FET5_1", "GPT2_4", "ILV5_4", "NRD1_1", "RDL1_4", "TFS1_1")

ggplot(
  data = filter(plates, prep_type == "+RT", target_id %in% target_id_levels_nice),
  aes(x = dilution, y = cq)
) +
  geom_point() +
  stat_smooth(
    formula = y ~ x, method = "lm", se = FALSE,
    aes(colour = "fit", linetype = "fit")
  ) +
  stat_smooth(
    formula = y ~ 1 + offset(-x * log(10) / log(2)), method = "lm", se = FALSE,
    aes(colour = "theory", linetype = "theory")
  ) +
  scale_x_log10(breaks = 10^-(0:3)) +
  scale_y_continuous(breaks = seq(0, 30, 2)) +
  labs(
    y = "Cycle count to threshold",
    title = "All reps, unnormalized",
    colour = "Dilution", linetype = "Dilution"
  ) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```


# Analyze amplification and melt curve data

## Load raw data for amplification and melt curves.


```{r load_amp,dependson="label_plates",results="show"}

plate1curve <- read_tsv("../inst/extdata/Edward_qPCR_Nrd1_calibration_2019-02-02.txt",
  skip = 2,
  col_names = c(
    "well", "SID", "Program", "Segment",
    "Cycle", "Time", "Temperature", "Fluor"
  )
) %>%
  debaseline() %>%
  left_join(plate1plan)

# amplification curve is program 2
platesamp <- plate1curve %>%
  filter(Program == 2)

# melt curve is program 3 or 4, depending on cycle
platesmelt <- plate1curve %>%
  filter(Program > 2) %>%
  calculate_drdt_plate() %>%
  filter(Temperature >= 61)
```


## Plot de-baseline'd raw data for single well

```{r plotamp_A1,dependson="load_amp",fig.width=4,fig.height=3}
ggplot(
  data = platesamp %>% filter(well == "A1"),
  aes(x = Cycle, y = Signal)
) +
  geom_line() +
  scale_y_continuous(expand = c(0.01, 0.01))
```

## Plot all amplification curves

Broken up by technical replicate here, to avoid overplotting.

```{r plotamp_all,dependson="load_amp",fig.height=11,fig.width=7}
ggplot(
  data = platesamp %>%
    filter(TechRep == "1"),
  aes(x = Cycle, y = Signal, colour = factor(dilution), linetype = prep_type)
) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  scale_linetype_manual(values = c("+RT" = "solid", "-RT" = "dashed", "NT" = "dotted")) +
  geom_line() +
  scale_x_continuous(breaks = seq(60, 100, 10), minor_breaks = seq(60, 100, 5)) +
  labs(title = "All Amp Curves, TechRep A")

ggplot(
  data = platesamp %>%
    filter(TechRep == "2"),
  aes(x = Cycle, y = Signal, colour = factor(dilution), linetype = prep_type)
) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  scale_linetype_manual(values = c("+RT" = "solid", "-RT" = "dashed", "NT" = "dotted")) +
  geom_line() +
  scale_x_continuous(breaks = seq(60, 100, 10), minor_breaks = seq(60, 100, 5)) +
  labs(title = "All Amp Curves, TechRep B")
```

## Plot melt curve for single well

```{r plotmelt_A1,dependson="load_amp",fig.width=4,fig.height=1.5}
ggplot(
  data = platesmelt %>%
    filter(well == "A1"),
  aes(x = Temperature, y = dRdT)
) +
  facet_wrap(~target_id) +
  geom_line() +
  scale_y_continuous(expand = c(0.02, 0.02))
```


## Plot all melt curves

Again broken up by technical replicate.

```{r plotmelt_all,dependson="load_amp",fig.height=11,fig.width=7}
ggplot(
  data = platesmelt %>%
    filter(TechRep == "1"),
  aes(x = Temperature, y = dRdT, colour = factor(dilution), linetype = prep_type)
) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  scale_linetype_manual(values = c("+RT" = "solid", "-RT" = "dashed", "NT" = "dotted")) +
  geom_line() +
  scale_x_continuous(breaks = seq(60, 100, 10), minor_breaks = seq(60, 100, 5)) +
  labs(title = "All Melt Curves, TechRep A")

ggplot(
  data = platesmelt %>%
    filter(TechRep == "2"),
  aes(x = Temperature, y = dRdT, colour = factor(dilution), linetype = prep_type)
) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  scale_linetype_manual(values = c("+RT" = "solid", "-RT" = "dashed", "NT" = "dotted")) +
  geom_line() +
  scale_x_continuous(breaks = seq(60, 100, 10), minor_breaks = seq(60, 100, 5)) +
  labs(title = "All Melt Curves, TechRep B")
```


## Plot zoomed melt curves

```{r plotmelt_zoomed,dependson="load_amp",fig.height=11,fig.width=7}
ggplot(
  data = platesmelt %>%
    filter(TechRep == "1", prep_type == "+RT"),
  aes(x = Temperature, y = dRdT, colour = factor(dilution))
) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  geom_line() +
  scale_x_continuous(
    breaks = seq(60, 100, 5), minor_breaks = seq(60, 100, 1),
    limits = c(73, 87)
  ) +
  labs(title = "Melt curves, zoomed, TechRep A") +
  theme(
    panel.grid.major.x = element_line(colour = "grey50", size = 0.4),
    panel.grid.minor.x = element_line(colour = "grey70", size = 0.1)
  )

ggplot(
  data = platesmelt %>%
    filter(TechRep == "2", prep_type == "+RT"),
  aes(x = Temperature, y = dRdT, colour = factor(dilution))
) +
  facet_grid(target_id ~ BioRep, scales = "free_y") +
  geom_line() +
  scale_x_continuous(
    breaks = seq(60, 100, 5), minor_breaks = seq(60, 100, 1),
    limits = c(73, 87)
  ) +
  labs(title = "Melt curves, zoomed, TechRep B") +
  theme(
    panel.grid.major.x = element_line(colour = "grey50", size = 0.4),
    panel.grid.minor.x = element_line(colour = "grey70", size = 0.1)
  )
```


## Plot only zoomed melt curves for nice probes

```{r plotmelt_zoomed_nice,dependson="load_amp",fig.height=6,fig.width=4}
target_id_levels_nice <- c("ECM38_3", "FET5_1", "GPT2_4", "ILV5_4", "NRD1_1", "RDL1_4", "TFS1_1")
ggplot(
  data = platesmelt %>%
    filter(
      TechRep == "1", prep_type == "+RT", dilution_nice == "1x",
      target_id %in% target_id_levels_nice
    ),
  aes(x = Temperature, y = dRdT, colour = BioRep)
) +
  facet_grid(target_id ~ ., scales = "free_y") +
  geom_line() +
  scale_x_continuous(
    breaks = seq(60, 100, 5), minor_breaks = seq(60, 100, 1),
    limits = c(73, 87)
  ) +
  labs(title = "Nice probes, TechRep A") +
  theme(
    panel.grid.major.x = element_line(colour = "grey50", size = 0.4),
    panel.grid.minor.x = element_line(colour = "grey70", size = 0.1)
  )
```


# Conclude acceptable primer sets

These probes have sensible standard curves, amplification curves, melt curves. In tie-breakers we pick the more highly detected probe.

* ECM38 set 3
* FET5 set 1 or 4
* GPT2 set 4
* ILV5 set 4
* NRD1 set 1 or 2
* RDL1 set 4
* TFS1 set 1
