---
title: "Calibration vignette"
author: "Edward Wallace"
date: "2 Feb 2019"
output:
  html_document:
    toc: true
    toc_depth: 2
---


# Plan

This vignette shows how to use tidyqpcr functions to calibrate qPCR probes

This is real qPCR data by Edward Wallace in Feb 2019, testing new RT-qPCR primer sets against _S. cerevisiae_ genes. We took exponential-phase total RNA previously extracted by Jamie Auxillos.

We tested 2-3 primer sets each for 7 genes with 2 Technical replicates as well:

* ECM38/ YLR299W (3 primer sets)
* FET5 / YFL041W (3 primer sets)
* GPT2/YKR067W
* ILV5/YLR355C
* NRD1/YNL251C (ordered 3rd primer set not tested today)
* RDL1/YOR285W
* TFS1/YLR178C

For each we do serial 5x dilutions down to 125x, a -RT and no template control. In duplicate.

## Conclude acceptable primer sets

These probes have sensible standard curves, amplification curves, melt curves. In tie-breakers we pick the more highly detected probe.

* ECM38 set 3
* FET5 set 1 or 4
* GPT2 set 4
* ILV5 set 4
* NRD1 set 1 or 2
* RDL1 set 4
* TFS1 set 1



```{r setup,warning=FALSE,message=FALSE,echo=FALSE}

## knitr options for report generation
knitr::opts_chunk$set(warning=FALSE,message=FALSE,echo=FALSE,cache=FALSE,
                      results="show")


## data processing options, packages, functions
options(stringsAsFactors=FALSE)  # load character strings as strings

library(reshape2) # data manipulation
library(tidyverse)
# 
library(readxl) # read excel spreadsheets

library(ggplot2) # plot graphs
library(cowplot) # extensions and nice defaults

 # set default theme for graphics
theme_set(theme_cowplot(font_size=11) %+replace% 
              theme(panel.border=element_rect(colour = "grey50",
                                            linetype = "solid",size=0.5),
                    strip.background = element_blank()))


scale_y_log2nice <- function(name=waiver(),omag=seq(-6,6),scilabels=FALSE,...) {
    # plot y axis on log2-scale with nice defaults
    breaks2 <- 2^omag
    if (scilabels) {
        labels2 <- paste("2^{",omag,"}",sep="")
    } else {
        labels2 <- breaks2
    }
    scale_y_log10(name,breaks=breaks2,labels=parse(text=labels2),...)
}

scale_y_log10nice <- function(name=waiver(),omag=seq(-6,6),scilabels=FALSE,...) {
    # plot y axis on log10-scale with nice defaults
    breaks10 <- 10^omag
    if (scilabels) {
        labels10 <- paste("10^{",omag,"}",sep="")
    } else {
        labels10 <- breaks10
    }
    scale_y_log10(name,breaks=breaks10,labels=parse(text=labels10),...)
}
```




```{r plate_functions}

create.blank.plate <- function(WellR=LETTERS[1:16],WellC=1:24) {
    ## create blank plate data frame
    plate <- expand.grid(WellR=WellR,WellC=WellC)
    plate$Well   <- with(plate,paste0(WellR,WellC))
    # plate$Sample <- NA
    # plate$Probe  <- NA
    return(plate)
}

label.plate.rowcol <- function(plate,rowkey=NULL,colkey=NULL) {
    ## label a plate by row and column keys
    if (!is.null(colkey)) {
        plate <- merge(plate,colkey,by="WellC")
    }
    if (!is.null(rowkey)) {
        plate <- merge(plate,rowkey,by="WellR")
    }
    return(plate[order(plate$WellR,plate$WellC),])
}

display_plate <- function(plate) {
    ggplot(data=plate,aes(x=factor(WellC),
                          y=factor(WellR,levels=rev(LETTERS)))) +
        geom_tile(aes(fill=Probe),alpha=0.3) +
        geom_text(aes(label=paste(Probe,Sample,Type,sep="\n")),size=2.5,lineheight=1) +
        scale_x_discrete(expand=c(0,0)) +
        scale_y_discrete(expand=c(0,0)) +
        coord_equal() +
        theme_void() + 
        theme(axis.text=element_text(angle=0),
              panel.grid.major=element_blank(),
              legend.position="none",
              plot.margin=unit(rep(0.01,4),"npc"),
              panel.border = element_blank())
}


getNormCt <- function(df,mycolumn="Ct",normProbes="ALG9",probename="Probe") {
    ### function to take data frame and attach a column to normalize things by.
    
    # make subset of df where gene is one of normGenes
    subdf <- df[df[[probename]] %in% normProbes,]
    
    # assign median of mycolumn to df$normct
    # note this is the same value for every row, a waste of space technically
    df$norm.by <- median(subdf[[mycolumn]],na.rm=TRUE)
    return(df)
}


normalizeqPCR <- function(df,mycolumn="Ct",normProbes="ALG9",probename="Probe") {
    # make normed count, grouped by Sample (biological rep)
    dfout <- 
        group_by(df,Sample) %>% # group by Sample
        do(getNormCt(.,mycolumn,normProbes,probename)) %>%      # get norm value for each Sample
        ungroup()                 # combine/ungroup again
    
    # Assign normalized values by dividing by normby
    dfout$Value.norm <- dfout[[mycolumn]] - dfout$norm.by
    dfout$Value.normexp <- 2^-dfout$Value.norm
                             
    return(dfout)
}

debaseline <- function(plateamp,maxcycle=10) {
    baseline <- 
        plateamp %>%
        group_by(Well) %>%
        filter(Program == 2, Cycle <= maxcycle) %>%
        summarize(Base = median(Fluor))
    plateamp %>% 
        left_join(baseline) %>%
        mutate(Signal=Fluor-Base)
        
}

getdRdT <- function(TT,RR,method=c("spline","diff"),...) {
    if (method == "diff") {
       return( -c(diff(RR)/diff(TT),NA) )
    } else if (method == "spline") {
        fit <- smooth.spline(x = TT, y=RR,...)
        return(-1 * predict(object = fit, x = TT, deriv = 1)$y )
    }
}

demelt <- function(platemelt) {
    platemelt %>%
        arrange(Well,Temperature) %>%
        group_by(Well) %>%
        mutate(dRdT=getdRdT(Temperature,Signal)) %>%
        ungroup()
}

```



## Label and plan plates



```{r label_plates,dependson="plate_functions"}
Names <- c("ECM38","FET5","GPT2","ILV5","NRD1","RDL1","TFS1")
Targets  <- c("YLR299W","YFL041W","YKR067W","YLR355C","YNL251C","YOR285W","YLR178C")
Namesrep <- c(rep(Names[1:2],each=3),rep(Names[3:7],each=2))
Probes <- paste(Namesrep,
                c(1,2,3,1,3,4,1,4,1,4,1,2,4,5,1,5),sep="_")

Probelevels <- Probes


rowkey <- data.frame(WellR=LETTERS[1:16],
                     Names=Namesrep,
                     # Targets=rep(Targets,each=2),
                     Probe=Probes
)

colkey <- data.frame(WellC=1:24,
                     Dilution=rep(c(1,1/5,1/25,1/125,1,1)),
                     DilutionNice=rep(c("1x","5x","25x","125x","-RT","NT")),
                     Type=c(rep("+RT",4),"-RT","NT"),
                     BioRep=rep(c("A","B"),each=12),
                     TechRep = rep(c("A","B"),each=6),
                     RT="SSIV")

plate1plan <- 
    label.plate.rowcol(
        create.blank.plate(),
        rowkey,colkey) %>%
    mutate(Sample=paste(BioRep,str_sub(RT,end=1L),DilutionNice,sep="_"))


```



```{r display_plates,fig.height=8,fig.width=12,dependson="label_plates"}
display_plate(plate1plan) 

```


## Load and summarize data

```{r load_plates,dependson="label_plates",results="show"}

# read my plates

plates <- read_tsv("../data/Edward qPCR Nrd1 calibration 2019-02-02 Ct.txt",
                       skip=1) %>%
    mutate(Well=Pos,Ct=Cp) %>%
    right_join(plate1plan)

```


```{r show_plates,dependson="load_plates",results="show"}
plates

summary(plates)
```

# Plot unnormalized data

## -RT and NT controls are low


```{r plot_unnormalized,dependson="load_plates",fig.height=6,fig.width=9}

ggplot(data=plates) +
    geom_point(aes(x=Probe,y=Ct,colour=DilutionNice,shape=Type),
               position=position_jitter(width = 0.2,height=0)) +
    labs(y="Cycle count to threshold",
         title="All reps, unnormalized") +
    facet_grid(RT~BioRep) +
    panel_border() +
    theme(axis.text.x=element_text(angle=90,vjust=0.5))
```



## Dilution series is linear - only used Superscript IV.


```{r plot_dilutions,dependson="load_plates",fig.height=11,fig.width=6}
ggplot(data=filter(plates,Type=="+RT"),aes(x=Dilution,y=Ct)) +
    geom_point() +
    stat_smooth(formula=y~x,method="lm",se=FALSE,
                aes(colour="fit",linetype="fit")) + 
    stat_smooth(formula=y~1+offset(-x*log(10)/log(2)),method="lm",se=FALSE,
                aes(colour="theory",linetype="theory")) + 
    scale_x_log10(breaks=10^-(0:3)) +
    scale_y_continuous(breaks=seq(0,30,2)) + 
    labs(y="Cycle count to threshold",
         title="All reps, unnormalized",
         colour="Dilution") +
    facet_grid(Probe~BioRep,scales="free_y") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5))
```

## Dilution series for nice probes only

```{r plot_dilutions_nice,dependson="load_plates",fig.height=6,fig.width=4}
Probesnice <- c("ECM38_3","FET5_1","GPT2_4","ILV5_4","NRD1_1","RDL1_4","TFS1_1")

ggplot(data=filter(plates,Type=="+RT",Probe %in% Probesnice),
       aes(x=Dilution,y=Ct)) +
    geom_point() +
    stat_smooth(formula=y~x,method="lm",se=FALSE,
                aes(colour="fit",linetype="fit")) + 
    stat_smooth(formula=y~1+offset(-x*log(10)/log(2)),method="lm",se=FALSE,
                aes(colour="theory",linetype="theory")) + 
    scale_x_log10(breaks=10^-(0:3)) +
    scale_y_continuous(breaks=seq(0,30,2)) + 
    labs(y="Cycle count to threshold",
         title="All reps, unnormalized",
         colour="Dilution") +
    facet_grid(Probe~BioRep,scales="free_y") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5))
```


## Load raw data for amplification and curves.


```{r load_amp,dependson="label_plates",results="show"}

plate1curve <- read_tsv("../data/Edward qPCR Nrd1 calibration 2019-02-02.txt",
                       skip=2,
                        col_names=c("Well","SID","Program","Segment",
                                    "Cycle","Time","Temperature","Fluor") 
                      ) %>%
    debaseline() %>%
    left_join(plate1plan) 


platesamp  <- plate1curve %>% filter(Program == 2)

platesmelt <- plate1curve %>% filter(Program > 2) %>% demelt() %>% filter(Temperature >= 61)

```


### Plot de-baseline'd raw data for single well

```{r testplottraj,dependson="load_amp",fig.width=4,fig.height=3}
ggplot(data=platesamp %>% filter(Well=="A1"),
       aes(x=Cycle,y=Signal)) + 
    geom_line() + 
    scale_y_log10()
```

### Plot all amplification curves

```{r print_techreps,results="show",echo=TRUE,cache=FALSE,eval="FALSE"}
plate1plan %>% 
           filter(TechRep=="A",Probe==Probes[1],DilutionNice=="1x")

plate1plan %>% 
           filter(TechRep=="B",Probe==Probes[1],DilutionNice=="1x")
```


```{r plotamp_all,dependson="load_amp",fig.height=11,fig.width=7}
ggplot(data=platesamp %>% 
           filter(TechRep=="A"),
       aes(x=Cycle,y=Signal,colour=factor(Dilution),linetype=Type)) + 
    facet_grid(Probe~RT+BioRep,scales="free_y") + 
    scale_linetype_manual(values=c("+RT"="solid","-RT"="dashed","NT"="dotted")) + 
    geom_line() +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="All Amp Curves, TechRep A") 

ggplot(data=platesamp %>% 
           filter(TechRep=="B"),
       aes(x=Cycle,y=Signal,colour=factor(Dilution),linetype=Type)) + 
    facet_grid(Probe~RT+BioRep,scales="free_y") + 
    scale_linetype_manual(values=c("+RT"="solid","-RT"="dashed","NT"="dotted")) + 
    geom_line() +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="All Amp Curves, TechRep B") 

```

Here it would be sensible to also estimate the probe efficiency.

## Plot melt curve

### Plot melt curve for single well

```{r plotmelt_A1,dependson="load_amp",fig.width=4,fig.height=1.5,eval=FALSE}
ggplot(data=platesmelt %>% 
           filter(Well=="A1"),
       aes(x=Temperature,y=dRdT)) + 
    facet_wrap(~Probe) + 
    geom_line() 

```


### Plot all melt curves

```{r plotmelt_all,dependson="load_amp",fig.height=11,fig.width=7}
ggplot(data=platesmelt %>% 
           filter(TechRep=="A"),
       aes(x=Temperature,y=dRdT,colour=factor(Dilution),linetype=Type)) + 
    facet_grid(Probe~RT+BioRep,scales="free_y") + 
    scale_linetype_manual(values=c("+RT"="solid","-RT"="dashed","NT"="dotted")) + 
    geom_line() +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="All Melt Curves, TechRep A") 

ggplot(data=platesmelt %>% 
           filter(TechRep=="B"),
       aes(x=Temperature,y=dRdT,colour=factor(Dilution),linetype=Type)) + 
    facet_grid(Probe~BioRep,scales="free_y") + 
    scale_linetype_manual(values=c("+RT"="solid","-RT"="dashed","NT"="dotted")) + 
    geom_line() +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="All Melt Curves, TechRep B") 

```


### Plot only zoomed melt curves for Superscript

```{r plotmelt_SS_zoomed,dependson="load_amp",fig.height=11,fig.width=7}
ggplot(data=platesmelt %>% 
           filter(TechRep=="A",Type=="+RT"),
       aes(x=Temperature,y=dRdT,colour=factor(Dilution))) + 
    facet_grid(Probe~BioRep,scales="free_y") + 
    geom_line() +
    scale_x_continuous(breaks=seq(60,100,5),minor_breaks=seq(60,100,1),
                       limits=c(73,87)) + 
    labs(title="Melt curves, zoomed, Superscript, TechRep A") +
    theme(panel.grid.major.x=element_line(colour="grey50",size=0.4),
          panel.grid.minor.x=element_line(colour="grey70",size=0.1))

ggplot(data=platesmelt %>% 
           filter(TechRep=="B",Type=="+RT",RT=="SSIV"),
       aes(x=Temperature,y=dRdT,colour=factor(Dilution))) + 
    facet_grid(Probe~RT+BioRep,scales="free_y") + 
    geom_line() +
    scale_x_continuous(breaks=seq(60,100,5),minor_breaks=seq(60,100,1),
                       limits=c(73,87)) + 
    labs(title="Melt curves, zoomed, Superscript, TechRep B") +
    theme(panel.grid.major.x=element_line(colour="grey50",size=0.4),
          panel.grid.minor.x=element_line(colour="grey70",size=0.1))

```


### Plot only zoomed melt curves for nice probes

```{r plotmelt_SS_zoomed_nice,dependson="load_amp",fig.height=6,fig.width=4}
Probesnice <- c("ECM38_3","FET5_1","GPT2_4","ILV5_4","NRD1_1","RDL1_4","TFS1_1")
ggplot(data=platesmelt %>% 
           filter(TechRep=="A",Type=="+RT",DilutionNice=="1x",
                  Probe %in% Probesnice),
       aes(x=Temperature,y=dRdT,colour=BioRep)) + 
    facet_grid(Probe~.,scales="free_y") + 
    geom_line() +
    scale_x_continuous(breaks=seq(60,100,5),minor_breaks=seq(60,100,1),
                       limits=c(73,87)) + 
    labs(title="Melt curves, zoomed, Nice probes, TechRep A") +
    theme(panel.grid.major.x=element_line(colour="grey50",size=0.4),
          panel.grid.minor.x=element_line(colour="grey70",size=0.1))
```

