---
title: "Multifactorial multi-plate qPCR analysis"
author: "Edward Wallace"
date: "June 2018"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Plan

This vignette shows how to use tidyqpcr functions to normalize and plot data from multifactorial experiments: many primer sets, many conditions, two plates.

This is real RT-qPCR data by Edward Wallace in June 2018, testing the effect of heat shock and transcription-targeting drugs in _Saccharomyces cerevisiae_ yeast.


## Pilot experiment

Do standard transcriptional inhibitors phenanthroline and thiolutin block the transcriptional heat shock response in yeast? This is a genuine question because some papers that argue that phenanthroline and thiolutin **induce** the transcriptional heat shock response.

Measure 16 primer sets: HOR7, HSP12, HSP26, HSP78, HSP104, RTC3, SSA4, PGK1, ALG9,  HHT2, HTB2, RPS3, RPS13, RPS15, RPS30A, RPL39.

Test 6 conditions. That's 3 transcriptional inhibitors (no drug control, 150ug/mL 1,10-phenanthroline, 3ug/mL thiolutin) in each of 2 conditions (- heat shock control, + heat shock 42C 10min), 2 biol reps each:

    - C- Control -HS
    - P- phenanthroline -HS
    - T- thiolutin -HS
    - C+ Control +HS
    - P+ phenanthroline +HS
    - T+ thiolutin +HS


```{r setup,warning=FALSE,message=FALSE,echo=FALSE}

## knitr options for report generation
knitr::opts_chunk$set(warning=FALSE,message=FALSE,echo=FALSE,cache=FALSE,
                      results="show")
library(tidyverse)
library(cowplot)
library(tidyqpcr)

 # set default theme for graphics
theme_set(theme_cowplot(font_size=11) %+replace% 
              theme(panel.border=element_rect(colour = "grey50",
                                            linetype = "solid",size=0.5),
                    strip.background = element_blank()))
```



## Label and plan plates


Reverse transcription by random primers (mixed with oligo-dT??)


```{r label_plates,dependson="plate_functions"}
Probes <- c("HOR7", "HSP12", "HSP26", "HSP78", "HSP104", "RTC3", "SSA4", "PGK1", "ALG9", " HHT2", "HTB2", "RPS3", "RPS13", "RPS15", "RPS30A", "RPL39")

HSlevels=c("-","+")
Druglevels=c("C","P","T")
Conditionlevels=paste0(Druglevels,rep(HSlevels,each=3))

rowkey <- data.frame(WellR=LETTERS[1:16],
                     Probe=factor(Probes,levels=Probes)
                     ) 

colkey <- data.frame(WellC=1:24,
                     Type=c(rep("+RT",18),rep("-RT",6)),
                     TechRep=rep(c(1,2,3,1),each=6) %>% factor(),
                     HS=factor(rep(HSlevels,each=3),levels=HSlevels),
                     Drug=factor(Druglevels,levels=Druglevels)
                     ) %>%
    mutate(Condition=paste0(Drug,HS) %>% factor(levels=Conditionlevels))

plateplan <-     
    label_plate_rowcol(create_blank_plate(WellR = LETTERS[1:16],WellC=1:24),
                       rowkey,colkey)

```



```{r display_plates,fig.height=9,fig.width=10,dependson="label_plates"}
display_plate( plateplan %>%
    mutate(Sample=paste0(Drug,HS)) )
```

## Load and summarize data

```{r load_plates,dependson="label_plates",results="show"}
# set working directory, if needed
# setwd("~/Dropbox (Drummond Lab)/EdinburghLab/RT-qPCR/2018-06/")
# read my plates

plate1 <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate1_Ct.txt",skip=1) %>%
    mutate(Well=Pos,Ct=Cp) %>%
    left_join(plateplan) %>%
    mutate(BiolRep="1",Plate="1")


plate2 <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate2_Ct.txt",skip=1) %>%
    mutate(Well=Pos,Ct=Cp) %>%
    left_join(plateplan) %>%
    mutate(BiolRep="2",Plate="2")

plates <- bind_rows(plate1,plate2) %>%
    mutate( Sample=paste0(Condition,BiolRep) )

summary(plates)

```

# Plot unnormalized data

## -RT controls are low



```{r plot_unnormalized,dependson="load_plates",fig.height=6,fig.width=8}
ggplot(data=plates) +
    geom_point(aes(x=Probe,y=Ct,shape=Condition,colour=Condition),
               position=position_jitter(width = 0.2,height=0)) +
    labs(y="Cycle count to threshold",
         title="All reps, unnormalized") +
    scale_shape_manual(values=c(15:18,5:6)) + 
    facet_grid(BiolRep~Type) +
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```


## Normalize Ct to PGK1, within Sample only 



```{r normalize_counts,dependson="load_plates"}
platesnorm <- plates %>% 
    filter(Type=="+RT") %>%
    normalizeqPCR(normProbes = "PGK1")

platesmed <- platesnorm %>%
    group_by(Sample,Condition,BiolRep,HS,Drug,Sample,Probe) %>%
    summarize(Ct=median(Value.norm,na.rm=TRUE),
              Abund=median(Value.normexp,na.rm=TRUE))

filter(platesmed,Probe=="HSP26")

```

## Plot normalized data, all reps

```{r plot_normalized,dependson="normalize_counts",fig.height=6,fig.width=6}
ggplot(data=platesnorm) +
    geom_point(aes(x=Probe,y=Value.norm,shape=Condition,colour=Condition),
               position=position_jitter(width = 0.2,height=0)) +
    labs(y="Ct relative to PGK1") +
    scale_shape_manual(values=c(15:18,5:6)) + 
    facet_grid(BiolRep~.) +
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```

## Plot normalized data, summarized vs Probe

```{r plot_normalizedsummarized1,dependson="normalize_counts",fig.height=3,fig.width=4}
ggplot(data=platesmed) +
    geom_point(aes(x=Probe,y=Abund,shape=BiolRep,colour=Condition),
               position=position_jitter(width = 0.2,height=0)) +
    scale_shape_manual(values=c(15:18,5:6)) + 
    scale_y_log10nice("mRNA relative detection") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```

```{r plot_normalizedsummarizedbyHS,dependson="normalize_counts",fig.height=3,fig.width=9}
ggplot(data=platesmed) +
    geom_point(aes(x=Probe,y=Abund,shape=BiolRep,colour=HS),
               position=position_jitter(width = 0.2,height=0)) +
    facet_wrap(~Drug,ncol=3) +
    scale_colour_manual(values=c("-"="grey50","+"="red")) +
    scale_y_log10nice("mRNA relative detection") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```

```{r plot_normalizedsummarizedbyDrug,dependson="normalize_counts",fig.height=3,fig.width=9}
ggplot(data=platesmed) +
    geom_point(aes(x=Probe,y=Abund,shape=BiolRep,colour=Drug),
               position=position_jitter(width = 0.2,height=0)) +
    facet_wrap(~HS,ncol=3) +
    scale_y_log10nice("mRNA relative detection") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```


# Melt and Amp Curves


```{r load_amp,dependson="label_plates",results="show"}

plate1curve <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate1.txt",
                       skip=2,
                        col_names=c("Well","SID","Program","Segment",
                                    "Cycle","Time","Temperature","Fluor") 
                      ) %>%
    debaseline() %>%
    left_join(plateplan) %>%
    mutate(BiolRep=1,Plate=1)

plate2curve <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate2.txt",
                       skip=2,
                        col_names=c("Well","SID","Program","Segment",
                                    "Cycle","Time","Temperature","Fluor") 
                      ) %>%
    debaseline() %>%
    left_join(plateplan)  %>%
    mutate(BiolRep=2,Plate=2)

# platesamp  <- plate1curve %>% filter(Program == 2)

# plate1curve
# plate2curve

platesamp <- bind_rows(plate1curve,plate2curve) %>% 
    filter(Program == 2)

platesmelt <- bind_rows(plate1curve,plate2curve) %>% 
    filter(Program != 2) %>% 
    demelt() %>% 
    filter(Temperature >= 61)  

```

## Melt Curves, BiolRep 1

```{r plotmelt_Rep1,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesmelt %>% 
           filter(TechRep==1,BiolRep==1),
       aes(x=Temperature,y=dRdT,linetype=Type)) + 
    facet_grid(Condition~Probe) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="Melt curves, BiolRep 1, Techrep 1") + panel_border()
```

## Melt Curves, BiolRep 2

```{r plotmelt_Rep2,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesmelt %>% 
           filter(TechRep==1,BiolRep==2),
       aes(x=Temperature,y=dRdT,linetype=Type)) + 
    facet_grid(Condition~Probe) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="Melt curves, BiolRep 2, TechRep 1") + panel_border()
```


## Amp Curves, BiolRep 1

```{r plotamp_Rep1,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesamp %>% 
           filter(TechRep==1,BiolRep==1),
       aes(x=Cycle,y=Fluor,linetype=Type)) + 
    facet_grid(Condition~Probe) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    expand_limits(y=0) + 
    labs(title="Amp. curves, BiolRep 1, Techrep 1") + panel_border()
```

```{r plotamp_Rep2,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesamp %>% 
           filter(TechRep==1,BiolRep==2),
       aes(x=Cycle,y=Fluor,linetype=Type)) + 
    facet_grid(Condition~Probe) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    expand_limits(y=0) + 
    labs(title="Amp. curves, BiolRep 2, Techrep 1") + panel_border()
```

