---
title: "Multifactorial multi-plate qPCR analysis example"
author: "Edward Wallace"
date: "June 2018"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{MultifactorialExample}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Summary: an example multifactorial qPCR experiment.

This vignette shows how to use tidyqpcr functions to normalize and plot data from multifactorial experiments: many primer sets, many conditions, two plates.

This is real RT-qPCR data by Edward Wallace in June 2018, testing the effect of heat shock and transcription-targeting drugs in _Saccharomyces cerevisiae_ yeast.


## Pilot experiment

Do standard transcriptional inhibitors phenanthroline and thiolutin block the transcriptional heat shock response in yeast? This is a genuine question because some papers that argue that phenanthroline and thiolutin **induce** the transcriptional heat shock response.

Measure 16 primer sets: HOR7, HSP12, HSP26, HSP78, HSP104, RTC3, SSA4, PGK1, ALG9,  HHT2, HTB2, RPS3, RPS13, RPS15, RPS30A, RPL39.

Test 6 conditions. That's 3 transcriptional inhibitors (no drug control, 150ug/mL 1,10-phenanthroline, 3ug/mL thiolutin) in each of 2 conditions (- heat shock control, + heat shock 42C 10min), 2 biol reps each:

    - C- Control -HS
    - P- phenanthroline -HS
    - T- thiolutin -HS
    - C+ Control +HS
    - P+ phenanthroline +HS
    - T+ thiolutin +HS


```{r setup,warning=FALSE,message=FALSE,echo=FALSE}

## knitr options for report generation
knitr::opts_chunk$set(warning=FALSE,message=FALSE,echo=TRUE,cache=FALSE,
                      results="show")
library(tidyverse)
library(cowplot)
library(tidyqpcr)

 # set default theme for graphics
theme_set(theme_cowplot(font_size=11) %+replace% 
              theme(panel.border=element_rect(colour = "grey50",
                                            linetype = "solid",size=0.5),
                    strip.background = element_blank()))
```



## Label and plan plates


Reverse transcription by random primers mixed with oligo-dT.


```{r label_plates,dependson="plate_functions"}

# list Targets / Probes
TargetIDs <- c("HOR7", "HSP12", "HSP26", "HSP78", 
            "HSP104", "RTC3", "SSA4", "PGK1", 
            "ALG9", " HHT2", "HTB2", "RPS3", 
            "RPS13", "RPS15", "RPS30A", "RPL39")

rowkey <- tibble(WellR=LETTERS[1:16],
                 TargetID=factor(TargetIDs,levels=TargetIDs) 
                 ) 

# Set up experimental samples
HSlevels <- c("-","+")
HSvalues <- factor(rep(HSlevels,each=3),levels=HSlevels)
Druglevels <- c("C","P","T")
Drugvalues <- factor(rep(Druglevels,times=2),levels=Druglevels)
Conditionlevels <- paste0(Druglevels,rep(HSlevels,each=3))
Conditionvalues <- factor(Conditionlevels,levels=Conditionlevels)

colkey <- create_colkey_6in24(HS=HSvalues,
                              Drug=Drugvalues,
                              Condition=Conditionvalues)

plateplan <-     
    label_plate_rowcol(create_blank_plate(WellR = LETTERS[1:16],WellC=1:24),
                       rowkey,colkey)

```

Display the plate plan using display_plate. 

```{r display_plates,fig.height=9,fig.width=10,dependson="label_plates"}
display_plate( plateplan %>%
    mutate(SampleID=Condition) )
```

Note that display_plate requires a column called `SampleID`, which here we had to make from the `Condition` varilable using `mutate(SampleID=Condition)`. The reason for doing this is that we have replicate samples of the same condition in different plates, and so we assign the unique sample name for each replicate after loading the plates together using `unite` in the next code chunk.

## Load and summarize data

```{r load_plates,dependson="label_plates",results="show"}
# read my plates

plate1 <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate1_Ct.txt",skip=1) %>%
    mutate(Well=Pos,Cq=Cp) %>%
    left_join(plateplan) %>%
    mutate(BiolRep="1",Plate="1")


plate2 <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate2_Ct.txt",skip=1) %>%
    mutate(Well=Pos,Cq=Cp) %>%
    left_join(plateplan) %>%
    mutate(BiolRep="2",Plate="2")

plates <- bind_rows(plate1,plate2) %>%
    unite(SampleID,Condition,BiolRep,sep="",remove=FALSE)

summary(plates)

```

# Plot unnormalized data

## -RT controls are low



```{r plot_unnormalized,dependson="load_plates",fig.height=6,fig.width=8}
ggplot(data=plates) +
    geom_point(aes(x=TargetID,y=Cq,shape=Condition,colour=Condition),
               position=position_jitter(width = 0.2,height=0)) +
    labs(y="Cycle count to threshold",
         title="All reps, unnormalized") +
    scale_shape_manual(values=c(15:18,5:6)) + 
    facet_grid(BiolRep~Type) +
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```


## Normalize Cq to PGK1, within Sample only 



```{r normalize_counts,dependson="load_plates"}
platesnorm <- plates %>% 
    filter(Type=="+RT") %>%
    calculate_deltacq_bysampleid(normTargetIDs = "PGK1")

platesmed <- platesnorm %>%
    group_by(SampleID,Condition,BiolRep,HS,Drug,TargetID) %>%
    summarize(Cq=median(Value.norm,na.rm=TRUE),
              Abund=median(Value.normexp,na.rm=TRUE))

filter(platesmed,TargetID=="HSP26")

```

## Plot normalized data, all reps

```{r plot_normalized,dependson="normalize_counts",fig.height=6,fig.width=6}
ggplot(data=platesnorm) +
    geom_point(aes(x=TargetID,y=Value.norm,shape=Condition,colour=Condition),
               position=position_jitter(width = 0.2,height=0)) +
    labs(y="Cq relative to PGK1") +
    scale_shape_manual(values=c(15:18,5:6)) + 
    facet_grid(BiolRep~.) +
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```

## Plot normalized data, summarized vs TargetID

### All on same axes

This plot shows all the summarized data on the same axes, but it is hard to pick out the different conditions by eye.

```{r plot_normalizedsummarized1,dependson="normalize_counts",fig.height=3,fig.width=4}
ggplot(data=platesmed) +
    geom_point(aes(x=TargetID,y=Abund,shape=BiolRep,colour=Condition),
               position=position_jitter(width = 0.2,height=0)) +
    scale_shape_manual(values=c(15:18,5:6)) + 
    scale_y_log10nice("mRNA relative detection") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```

### Faceted by drug treatment

This plot shows all the summarized data "faceted" on different axes for different drug treatments. It highlights that, for example, SSA4 detection increases in response to heat in all drug treatments.

```{r plot_normalizedsummarizedbyHS,dependson="normalize_counts",fig.height=3,fig.width=9}
ggplot(data=platesmed) +
    geom_point(aes(x=TargetID,y=Abund,shape=BiolRep,colour=HS),
               position=position_jitter(width = 0.2,height=0)) +
    facet_wrap(~Drug,ncol=3) +
    scale_colour_manual(values=c("-"="grey50","+"="red")) +
    scale_y_log10nice("mRNA relative detection") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```

### Faceted by heat shock condition

By contrast, this plot shows all the summarized data "faceted" on different axes for different conditions. This shows that there is no clear response to the drug treatments in either condition.

```{r plot_normalizedsummarizedbyDrug,dependson="normalize_counts",fig.height=3,fig.width=9}
ggplot(data=platesmed) +
    geom_point(aes(x=TargetID,y=Abund,shape=BiolRep,colour=Drug),
               position=position_jitter(width = 0.2,height=0)) +
    facet_wrap(~HS,ncol=3) +
    scale_y_log10nice("mRNA relative detection") + 
    theme(axis.text.x=element_text(angle=90,vjust=0.5),
          panel.border=element_rect(fill = NA,linetype=1,
                                    colour = "grey50",size=0.5))

```


# Melt and Amp Curves

Note: this chunk throwing an error in CHECK.

```{r load_amp,dependson="label_plates",results="show"}

plate1curve <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate1.txt",
                       skip=2,
                        col_names=c("Well","SID","Program","Segment",
                                    "Cycle","Time","Temperature","Fluor") 
                      ) %>%
    debaseline() %>%
    left_join(plateplan) %>%
    mutate(BiolRep=1,Plate=1)

plate2curve <- read_tsv("../inst/extdata/Edward_qPCR_TxnInhibitors_HS_2018-06-15_plate2.txt",
                       skip=2,
                        col_names=c("Well","SID","Program","Segment",
                                    "Cycle","Time","Temperature","Fluor") 
                      ) %>%
    debaseline() %>%
    left_join(plateplan)  %>%
    mutate(BiolRep=2,Plate=2)

# platesamp  <- plate1curve %>% filter(Program == 2)

# plate1curve
# plate2curve

platesamp <- bind_rows(plate1curve,plate2curve) %>% 
    filter(Program == 2)

platesmelt <- bind_rows(plate1curve,plate2curve) %>% 
    filter(Program != 2) %>% 
    calculate_drdt_plate() %>% 
    filter(Temperature >= 61)  

```

## Melt Curves, BiolRep 1

```{r plotmelt_Rep1,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesmelt %>% 
           filter(TechRep==1,BiolRep==1),
       aes(x=Temperature,y=dRdT,linetype=Type)) + 
    facet_grid(Condition~TargetID) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="Melt curves, BiolRep 1, Techrep 1") + panel_border()
```

## Melt Curves, BiolRep 2

```{r plotmelt_Rep2,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesmelt %>% 
           filter(TechRep==1,BiolRep==2),
       aes(x=Temperature,y=dRdT,linetype=Type)) + 
    facet_grid(Condition~TargetID) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    scale_x_continuous(breaks=seq(60,100,10),minor_breaks=seq(60,100,5)) + 
    labs(title="Melt curves, BiolRep 2, TechRep 1") + panel_border()
```


## Amp Curves, BiolRep 1

```{r plotamp_Rep1,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesamp %>% 
           filter(TechRep==1,BiolRep==1),
       aes(x=Cycle,y=Signal,linetype=Type)) + 
    facet_grid(Condition~TargetID) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    expand_limits(y=0) + 
    labs(title="Amp. curves, BiolRep 1, Techrep 1") + panel_border()
```

```{r plotamp_Rep2,dependson="load_amp",fig.width=12,fig.height=6}
ggplot(data=platesamp %>% 
           filter(TechRep==1,BiolRep==2),
       aes(x=Cycle,y=Signal,linetype=Type)) + 
    facet_grid(Condition~TargetID) + 
    geom_line() +
    scale_linetype_manual(values=c("-RT"="dashed","+RT"="solid")) +
    expand_limits(y=0) + 
    labs(title="Amp. curves, BiolRep 2, Techrep 1") + panel_border()
```

